### 2017-07-22 chapters/introduction/mechanics.tex
Add latest

\subsection{Kinematics}
We represent the heart as a continuum body $\mathfrak{B}$ embedded in
$\mathbb{R}^3$. A configuration of $\mathfrak{B}$ is a mapping $\chi:
\mathfrak{B} \rightarrow \mathbb{R}^3$. 
We denote the \emph{reference configuration} of the heart by $\Omega
\equiv \chi_0(\mathfrak{B})$, and the \emph{current configuration} by $\omega
\equiv \chi(\mathfrak{B})$. The mapping $\varphi :  \Omega
\rightarrow \omega$ given by the composition $\varphi = \chi
\circ \chi_0^{-1}$ is a smooth, orientation preserving (positive
determinant) and invertible map. We denote the coordinates in the
reference configuration by $\Xvec \in \Omega$, and the coordinates in the current
configuration by $\xvec \in \omega$. The coordinates $\Xvec$ and $\xvec$ are
commonly referred to as material and spatial points respectively, and
are related through the mapping $\varphi$, by $\xvec = \varphi(\Xvec)$.
For time-dependent problems it is common to make  the time-dependence
explicitly by writing $\xvec = \varphi(\Xvec, t)$. In the following
we will only focus on the mapping between two configurations and
therefore no time-dependence is needed. The mapping $\varphi$ will be
referred to as the \emph{motion}. The \emph{deformation gradient} is a
rank-2 tensor, defined as the partial derivative of the motion with
respect to the material coordinates:
\begin{align}
  \F(\Xvec) = \frac{\partial \varphi}{\partial \Xvec} = \Grad \xvec.
  \label{eq:deformation_gradient}
\end{align}
Here we also introduce the notation $\Grad$, which means derivative
with respect to reference coordinates.
The deformation gradient maps vectors in the reference configuration to
vectors in the current configuration, and belongs to the space of
linear transformations from $\mathbb{R}^3$ to $\mathbb{R}^3$ with
strictly positive determinant, which we denote by
$\mathrm{Lin}^+$. Another important quantity is the
\emph{displacement} field  
\begin{align}
  \uvec = \xvec-\Xvec, 
  \label{eq:displacement}
\end{align}
which relates positions in the reference configuration to positions
in the current configuration. From \eqref{eq:deformation_gradient} we
see that
\begin{align}
  \F = \Grad \xvec = \Grad \uvec + \Grad \Xvec = \Grad \uvec + \I
\end{align}
Some other useful quantities are the \emph{right Cauchy-Green} deformation
tensor $\C = \F^T\F$, the \emph{left Cauchy-Green} deformation tensor
$\mathbf{B} = \F\F^T$, the \emph{Green-Lagrange} strain tensor
$\mathbf{E} = \frac{1}{2}(\C - \I)$, and the determinant of the
deformation gradient $J = \det \F$.

An important concept in mechanics is the concept of stress, which is
defined as force per area
$\left[\frac{\mathrm{N}}{\mathrm{m}^2}\right]$. When working with
different configurations one needs to be careful with which forces and
which areas we are talking about. Table \ref{tab:stress_tensor}
shows how forces and areas are related for the most important stress
tensors used in this thesis. Note that the explicit form of the stress
tensor requires a consitutive law for the material at hand. This will
be discussed in more detail in Section \ref{sec:constitutive_relations}.

@@

\subsection{Balance laws and transformations}
In this section we will cover some basic transformation used to
derrive the fore-balance equations for the mechanics of the heart.

\subsubsection{Transformations between reference and current
  configuration}
By definition the reference configuration, $\Omega$ and current
configuration $\omega$ are related via the motion $\varphi$ in the
sense that a point $\mathfrak{p}$ with reference coordinates $\Xvec$ and current
coordinates $\xvec$ satisfies $\xvec = \varphi(\Xvec)$. Likewise a
vector in the $\Omega$ is related to a vector in $\omega$ via the
deformation gradient $\F$; if $\mathrm{d}\Xvec$ is a vector in the
reference configuration it will transform to the vector
$\mathrm{d}\xvec$ in the current configuration, and $\mathrm{d}\xvec =
\F \mathrm{d}\Xvec$. From this relation we also derive that the
transformation of an infinitesimal volume element in the reference
configuration, $\mathrm{d}V$ is related to an infinitesimal volume
element in the current configuration, $\mathrm{d}v$  via the determinant of the
deformation gradient,
\begin{align}
  \mathrm{d}v =\det(\F) \mathrm{d}V.
  \label{eq:volume_element}
\end{align}
Another important transformation is the transformation of normal
vectors. By noting that we can write \eqref{eq:volume_element} using
surface elements
\begin{align*}
  \mathrm{d}s \mathbf{n} \mathrm{d}\xvec  &= \mathrm{d}v = \det(\F) \mathrm{d}V = \det(\F) \mathrm{d}S  \Nvec \mathrm{d}\Xvec\\
  &\implies \left( \mathrm{d}s \mathbf{n} \F  - \mathrm{d}S \det(\F) \Nvec \right) \mathrm{d}\Xvec = 0\\
  &\implies \left( \mathrm{d}s \F^T \mathbf{n}  - \mathrm{d}S \det(\F) \Nvec \right) \mathrm{d}\Xvec = 0,\\
\end{align*}
we get \emph{Nanson's formula}
\begin{align}
  \mathrm{d}s \mathbf{n}  =  \det(\F) \F^{-T} \mathrm{d}S \Nvec,
\end{align}
which relates the normal vector in the current configuration to the
normal vector in the reference configuration.

@@

\begin{align*}
  \int_{\partial \Omega} \mathbf{t} \mathrm{d}s
  = \int_{\partial \Omega} \sigma \mathbf{n} \mathrm{d}s
  = \int_{\Omega} \nabla \cdot \sigma \mathrm{d}v,
\end{align*}
and by collecting the terms from \eqref{eq:cons_lin_mom}. We arrive at
Cauchys momentum equation
\begin{align}
  \nabla \cdot \sigma + \mathbf{b} =  \rho \dot{\mathbf{v}}.
  \label{eq:chauch_momentum_eq}
\end{align}
The contribution for the body force ($\mathbf{b}$)  and inertial term
($\rho \dot{\mathbf{v}}$) are negligible compared to the stresses
\cite{hunter1996kd,tallarida1970left, moskowitz1981effects}, which is
why the force balance equations is typically only stated as
\begin{align}
  \nabla \cdot \sigma = \mathbf{0}.
  \label{eq:momentum_simple_current}
\end{align}
Note that we have formulated the balance law in the current
configuration. An equivalent statement can be formulated in terms of
the reference configuration
\begin{align}
  \nabla \cdot \FPK = \mathbf{0}, 
  \label{eq:momentum_simple_reference}
\end{align}
where $\FPK$ is the first Piola-Kirchhoff stress tensor.

@@

Even though experimental studies have indicated a visco-elastic behavior
of the myocardium \cite{dokos2002shear, gultekin2016orthotropic}, a
common assumption is to consider a quasi-static behavior and therefore
it is also possible to model the myocardium as hyperelastic.
This means that we postulate the existence of
a strain-energy density function $\Psi:\mathrm{Lin}^+ \rightarrow
\mathbb{R}^+$, and that stress is given by the relation
\begin{align}
\FPK = \frac{\partial \Psi(\F)}{\partial \F}.
\end{align}
Since stress has unit Pa, we see that the strain-energy density
function is defined as energy per unit reference volume, and has units
$\frac{\text{Joule}}{m^3}$. 
The strain-energy density function relates  the amount of
energy that is stored within the material in response to a given
strain. Hence, the stresses in a hyperelastic material with a given
strain-energy density function, depends only on the strain, and not the
path for which the material deforms. On the contrary, if the model had
been visco-elastic we would expect to see hysteresis in the
stress/strain curve, but this is not possible for a hyperelastic
material.

\begin{remark}
  The second law of thermodynamics states that the total entropy
  production in a thermodynamic process can never be nagative. Elastic
  materials defines a special class of materials in which the entropy
  production is zero. Within this thermodynamic framework the
  strain-energy density function coincides (up to a constant) with the
  Helmholtz free energy density.
\end{remark}

\subsubsection{General requirements for the strain-energy density
  function}
\label{sec:strain_energy_req}
Some general requirements must hold for the strain-energy function.
First of all, we require that the reference state is stress free and
that the stored energy increases monotonically with the deformation. 
Formally this can be stated simply as 
\begin{align*}
  \Psi(\I) = 0 && \Psi(\F) \geq 0.
\end{align*}
Moreover, if expanding or compressing a body to zero volume would
require an infinite amount of energy:
\begin{align*}
  \Psi(\F) \rightarrow \infty \; && \text{as} &&\; \det \F \rightarrow& 0 \\
  \Psi(\F) \rightarrow \infty \; && \text{as} &&\; \det \F \rightarrow& \infty
\end{align*}
We say that the strain energy should be objective, meaning that the
stored energy in the material should be invariant with respect to
change of observer. Formally we must have: \emph{given any positive symmetric
rank-2 tensor $\C \in \mathrm{Sym}$:}
\begin{align}
  \Psi(\mathbf{C}) = \Psi(\mathbf{Q}\mathbf{C}\mathbf{Q}^T), \; \forall \mathbf{Q} \in \mathcal{G} \subseteq \mathrm{Orth}.
\end{align}
Here $\mathrm{Orth}$ is the group of all positive orthogonal matrices.
If $\mathcal{G} = \mathrm{Orth}$ we say that the material is
isotropic, and otherwise we say that the material is anisotropic.
This brings us to another important issue, which is related to the
choice of coordinate-system. Having to deal with different
coordinate-systems, and mapping quantities from one coordinate-system
to another can results in complicated computations. Therefore it would be beneficial if we
could work with quantities which do not depend on the choice of
coordinate-system. Such quantities are called invariants. 
If the material is isotropic, the representation theorem for
invariants states that $\Psi$ can be expressed in terms of the
principle invariants of $\mathbf{C}$, that is $\Psi = \Psi(I_1, I_2,
I_3)$. The principle invariants $I_i, i=1,2,3$ are the coefficients in
the characteristic polynomial of $\mathbf{C}$, and is given by 
\begin{align}
  I_1 = \tr \C,  && I_2 = \frac{1}{2}\left[ I_1^2 - \tr(\C^2)\right] && \text{and} && I_3 = \det \C.
\end{align}
In the case when the material constitutes a transversely isotropic
behavior, that is the material has a preferred direction $\mathbf{a}_0$,
which in the case of the myocardium could be the direction of fiber
muscle fibers, we have
\begin{align*}
  \mathcal{G} = \{ \mathbf{Q} \in \mathrm{Orth}: \mathbf{Q}(\mathbf{a}_0\otimes\mathbf{a}_0)\mathbf{Q}^T\},
\end{align*}
with $\otimes$ being the outer product. In this case the strain-energy
density function can still be expressed through invariants. However,
we need to included the so called quasi-invariants, which are defined
as stretches in the local microstructural coordinate-system. The
transversely isotropic invariants is given by
\begin{align*}
  I_{4\mathbf{a}_0 } = \mathbf{a}_0 \cdot (\C \mathbf{a}_0) && \text{and} && I_5 = \mathbf{a}_0 \cdot (\C^2 \mathbf{a}_0).
\end{align*}
The invariants do have a physical interpretation. For instance, $I_3$
is related to the volume ratio of material during deformation, while
$I_{4\mathbf{a}_0 } $ is related to the stretch along the direction
$\mathbf{a}_0 $. Indeed the \emph{stretch} ratio in the direction
$\mathbf{a}_0$ is given by $\lambda_{\mathbf{a}_0} = | \F \mathbf{a}_0
|$ and we see that $I_{4\mathbf{a}_0 }  =  \mathbf{a}_0 \cdot (\C
\mathbf{a}_0) = \F \mathbf{a}_0 \cdot (\F \mathbf{a}_0) =
\lambda_{\mathbf{a}_0}^2$. For more details about invariants see
\cite{holzapfel2009constitutive,liu1982representations}.

The theory of global existence of unique solutions for elastic problem
was originally based convexity of the free energy function.
A function $\phi: \Omega \rightarrow \mathbb{R}$ is \emph{convex} if for any
$\Xvec_1, \Xvec_2 \in \Omega$ we have
\begin{align}
  \phi(\lambda \Xvec_1+ (1-\lambda) \Xvec_2)
  \leq \lambda \phi(\Xvec_1)
  + (1-\lambda) \phi(\Xvec_2).
\end{align}
However, from a physical point of view this requirement is to strict
\cite{ball1976convexity}. A slightly weaker requirement is that the
strain-energy function $\Psi$ is \emph{polyconvex}, meaning that there exist
a convex function $\phi$ such that
\begin{align*}
  \Psi(\F) = \phi(\F, \cof \F, \det \F).
\end{align*}
Note however, that if $\Psi$ is convex then it is also polyconvex.

@@

The myocardium is composed by small blood vessels that supply the
myocardial cells with oxygen. When the myocardium contracts this
perfused blood is then squeezed out, resulting in an overall loss of
2-4\% volume\cite{yin1996compressibility}. Such materials are
referred to as compressible. When the 
volume are preserved we refer to the material as incompressible.
Since 2-4\% is very little, a common assumption in cardiac modeling,
which has also been made in the work conducted in this thesis, is to assume
the myocardium to be incompresible. The reason for this choice is
purely numerical.

For an incompressible material, only isochoric motions are
possible. This means that volume of the material do not change during
any deformation and hence we have the constraint 
\begin{align}
  J = \det(\F) = 1.
  \label{eq:incompressible_cons}
\end{align}
The constraint \eqref{eq:incompressible_cons} can be imposed by
considering the modified strain energy function
\begin{align}
  \Psi = \Psi(\F) + p(J-1),
  \label{eq:incomp_strain_energy}
\end{align}
where $p$ is a scalar which serves as a Lagrange multiplier, but which
can be identified as the hydrostatic pressure. Like the displacement,
the hydrostatic pressure is unknown and has to be determined from the equilibrium
equations \eqref{eq:force_balance_strong_ref}. If we differentiate
\eqref{eq:incomp_strain_energy} with respect to $\F$ we get the First
Piola-Kirchhoff stress tensor for an incompressible material
\begin{align}
  \FPK = \frac{\partial \Psi(\F)}{\partial \F} + J p \F^{-T}.
\end{align}
Likewise the Cauchy stress tensor is given by
\begin{align}
  \Cauchy = J^{-1} \frac{\partial \Psi(\F)}{\partial \F}\F^{T} + p \I.
  \label{eq:cauchy_incomp}
\end{align}

@@

\subsubsection{Uncoupling of volumetric and isochoric response}
The total strain energy function in \eqref{eq:incomp_strain_energy}
is written as a sum of isochoric and volumetric components. Let
\begin{align}
  \F =  \F_{\mathrm{vol}} \F_{\mathrm{iso}},
\end{align}
then $ \F_{\mathrm{vol}} =
J^{1/3}\I$ and $\F_{\mathrm{iso}} = J^{-1/3}\F$. For
compressible material (i.e with $J \neq 1$) it is important to consider
only deviatoric strains in the strain-energy density function, so that
$\Psi = \Psi_{\mathrm{iso}}(\F_{\mathrm{iso}}) +
\Psi_{\mathrm{vol}}(J)$. For incompressible material ($J = 1$), we
have $\F_{\mathrm{vol}} = \I$ so that such a decomposition seems
uneccesary. However, a similar decomposition has shown to be
numerically beneficial \cite{weiss1996finite}. Note that, in this case, a similar
decoupling of the stress tensors has to be done.

@@

Choosing the correct boundary conditions for the model is essential,
and the choice should mimic what is observed in the reality. To
physiologically constrain the ventricle in a correct way is difficult,
and different approaches has been proposed.
The boundary condition at the endocardium is typically modeled as a
Neumann boundary condition, representing the endocardial blood
pressure, for the left ventricle we have
\begin{align}
  \sigma \mathbf{n} = -p_{\mathrm{lv}} \mathbf{n}, \;  \xvec \in  \lVendo(t), 
\end{align}
and for the right ventricle, ``lv'' is substituted with ``rv''.
This condition has a negative sign because the unit normal
$\mathbf{N}$ is pointing out of the domain, while the pressure is
acting into the domain. 
Note that this condition is imposed on the current configuration, and
to utilize the Lagrangian formulation we can pull back this condition
to the reference configuration to obtain
\begin{align}
  \FPK\mathbf{N} &= -p_{\mathrm{lv}} J\mathbf{F}^{-T} \cdot \mathbf{N}, \;  \Xvec \in \lVendo
\end{align}
Likewise, it is common to enforce a
Neumann boundary condition on the epicardium,
\begin{align}
\FPK\mathbf{N}  &= -p_{\mathrm{epi}}  J\mathbf{F}^{-T} \cdot \mathbf{N}, \;  \Xvec \in \epi.
\end{align}
However, the pressure $p_{\mathrm{lv}}$ is usually set
to zero.

@@

\subsection{Force-balance equation}
We will now collect all the terms that makes up to force balance for
the cardiac mechanics problem. Considering the myocardium as a incompressible,
hyperelastic material with a Robin type boundary condition at the base
we obtain the following set of equations in the Lagrangian formulation
\begin{align}
  \begin{split}
  \nabla \cdot \FPK &= 0 \\
  J - 1 &= 0,
  \end{split}
 \label{eq:force_balance_strong}
\end{align}
with $J = \det(\F)$, and appropriate boundary conditions.
% \begin{align}
%   \FPK \Nvec + k \uvec &= \mathbf{0},  \;  \Xvec \in \lvbase\\
%   u_1 &= 0,  \;  \Xvec \in \lvbase \\
%   \FPK\mathbf{N}  &= \mathbf{0}, \;  \Xvec \in \epi \\
%   \FPK\mathbf{N} &= -p_{\mathrm{lv}} J\mathbf{F}^{-T} \cdot \mathbf{N}, \;  \Xvec \in \lvendo
%   \label{eq:bndry_conditions_strong}
% \end{align}

@@

  \Pi(\mathbf{u}, p) &= \Pi_{\mathrm{int}}(\mathbf{u},p) + \Pi_{\mathrm{ext}}(\mathbf{u}). \\
  \Pi_{\mathrm{int}}(u,p) &= \int_{\Omega} \left[ p(J(\mathbf{u}) - 1) +  \Psi(\mathbf{F}(\mathbf{u})) \right] \mathrm{d}V\\

@@

>>>>

\subsection{Kinematics}
We represent the heart as a continuum body $\mathfrak{B}$ embedded in
$\mathbb{R}^3$. A configuration of $\mathfrak{B}$ is a mapping $\chi:
\mathfrak{B} \rightarrow \mathbb{R}^3$. 
We denote the \emph{reference configuration} of the heart by $\Omega
\equiv \chi_0(\mathfrak{B})$, and the \emph{current configuration} by $\omega
\equiv \chi(\mathfrak{B})$. The mapping $\varphi :  \Omega
\rightarrow \omega$, given by the composition $\varphi = \chi
\circ \chi_0^{-1}$, is a smooth, orientation preserving (positive
determinant) and invertible map. We denote the coordinates in the
reference configuration by $\Xvec \in \Omega$, and the coordinates in the current
configuration by $\xvec \in \omega$. The coordinates $\Xvec$ and $\xvec$ are
commonly referred to as material and spatial points respectively, and
are related through the mapping $\varphi$, by $\xvec = \varphi(\Xvec)$.
For time-dependent problems it is common to make  the time-dependence
explicitly by writing $\xvec = \varphi(\Xvec, t)$. In the following
we will only focus on the mapping between two configurations and
therefore no time-dependence is needed. The mapping $\varphi$ will be
referred to as the \emph{motion}. The \emph{deformation gradient} is a
rank-2 tensor, defined as the partial derivative of the motion with
respect to the material coordinates:
\begin{align}
  \F(\Xvec) = \frac{\partial \varphi}{\partial \Xvec} = \Grad \xvec.
  \label{eq:deformation_gradient}
\end{align}
Here we also introduce the notation $\Grad$, which means derivative
with respect to reference coordinates.
The deformation gradient maps vectors in the reference configuration to
vectors in the current configuration, and belongs to the space of
linear transformations from $\mathbb{R}^3$ to $\mathbb{R}^3$ with
strictly positive determinant, which we denote by
$\mathrm{Lin}^+$. Another important quantity is the
\emph{displacement} field  
\begin{align}
  \uvec = \xvec-\Xvec, 
  \label{eq:displacement}
\end{align}
which relates positions in the reference configuration to positions
in the current configuration. From \eqref{eq:deformation_gradient} we
see that
\begin{align}
  \F = \Grad \xvec = \Grad \uvec + \Grad \Xvec = \Grad \uvec + \I.
\end{align}
Some other useful quantities are the \emph{right Cauchy-Green} deformation
tensor $\C = \F^T\F$, the \emph{left Cauchy-Green} deformation tensor
$\mathbf{B} = \F\F^T$, the \emph{Green-Lagrange} strain tensor
$\mathbf{E} = \frac{1}{2}(\C - \I)$, and the determinant of the
deformation gradient $J = \det \F$.

An important concept in mechanics is the concept of stress, which is
defined as force per area
$\left[\frac{\mathrm{N}}{\mathrm{m}^2}\right]$. When working with
different configurations one needs to be careful with which forces and
which areas we are talking about. Table \ref{tab:stress_tensor}
shows how forces and areas are related for the most important stress
tensors used in this thesis. Note that the explicit form of the stress
tensor requires a constitutive law for the material at hand. This will
be discussed in more detail in Section \ref{sec:constitutive_relations}.

@@

\subsection{Balance laws and transformations}
In this section we will cover some basic transformation used to
derive the fore-balance equations for the mechanics of the heart.

\subsubsection{Transformations between reference and current
  configuration}
By definition, the reference configuration $\Omega$, and current
configuration $\omega$, are related via the motion $\varphi$ in the
sense that a point $\mathfrak{p} \in \mathfrak{B}$ with reference
coordinates $\Xvec$ and current coordinates $\xvec$ satisfies $\xvec =
\varphi(\Xvec)$. Likewise a vector in the reference configuration is
related to a vector in the current configuration  via the
deformation gradient $\F$; if $\mathrm{d}\Xvec$ is a vector in the
reference configuration it will transform to the vector
$\mathrm{d}\xvec$ in the current configuration, and $\mathrm{d}\xvec =
\F \mathrm{d}\Xvec$. From this relation we also derive that the
transformation of an infinitesimal volume element in the reference
configuration, $\mathrm{d}V$ is related to an infinitesimal volume
element in the current configuration, $\mathrm{d}v$  via the determinant of the
deformation gradient,
\begin{align}
  \mathrm{d}v =\det(\F) \mathrm{d}V.
  \label{eq:volume_element}
\end{align}
Another important transformation is the transformation of normal
vectors. By noting that we can write \eqref{eq:volume_element} using
surface elements
\begin{align*}
  \mathrm{d}s \mathbf{n} \mathrm{d}\xvec  &= \mathrm{d}v = \det(\F) \mathrm{d}V = \det(\F) \mathrm{d}S  \Nvec \mathrm{d}\Xvec\\
  &\implies \left( \mathrm{d}s \mathbf{n} \F  - \mathrm{d}S \det(\F) \Nvec \right) \mathrm{d}\Xvec = 0\\
  &\implies \left( \mathrm{d}s \F^T \mathbf{n}  - \mathrm{d}S \det(\F) \Nvec \right) \mathrm{d}\Xvec = 0,\\
\end{align*}
we get \emph{Nanson's formula}
\begin{align}
  \mathrm{d}s \mathbf{n}  =  \det(\F) \F^{-T} \mathrm{d}S \Nvec,
\end{align}
which relates the normal vector in the current configuration to the
normal vector in the reference configuration.

@@

\begin{align*}
  \int_{\partial \Omega} \mathbf{t} \mathrm{d}s
  = \int_{\partial \Omega} \sigma \mathbf{n} \mathrm{d}s
  = \int_{\Omega} \nabla \cdot \sigma \mathrm{d}v,
\end{align*}
and by collecting the terms from \eqref{eq:cons_lin_mom}. We arrive at
Cauchys momentum equation
\begin{align}
  \nabla \cdot \sigma + \mathbf{b} =  \rho \dot{\mathbf{v}}.
  \label{eq:chauch_momentum_eq}
\end{align}
The contribution from the body force ($\mathbf{b}$)  and inertial term
($\rho \dot{\mathbf{v}}$) are negligible compared to the stresses
\cite{hunter1996kd,tallarida1970left, moskowitz1981effects}, which is
why the force balance equations is typically only stated as
\begin{align}
  \nabla \cdot \sigma = \mathbf{0}.
  \label{eq:momentum_simple_current}
\end{align}
Note that we have formulated the balance law in the current
configuration. An equivalent statement can be formulated in terms of
the reference configuration
\begin{align}
  \nabla \cdot \FPK = \mathbf{0}, 
  \label{eq:momentum_simple_reference}
\end{align}
where $\FPK$ is the first Piola-Kirchhoff stress tensor.

@@

Even though experimental studies have indicated a visco-elastic behavior
of the myocardium \cite{dokos2002shear, gultekin2016orthotropic}, a
common assumption is to consider a quasi-static behavior, meaning that
the inertial term in \eqref{eq:chauch_momentum_eq} is negligible and
static equilibrium can be achieved at any time. Therefore
it is also possible to model the myocardium as a hyperelastic material
which is a type of elastic material.
This means that we postulate the existence of
a strain-energy density function $\Psi:\mathrm{Lin}^+ \rightarrow
\mathbb{R}^+$, and that stress is given by the relation
\begin{align}
\FPK = \frac{\partial \Psi(\F)}{\partial \F}.
\end{align}
Since stress has unit Pa, we see that the strain-energy density
function is defined as energy per unit reference volume, and has units
$\frac{\text{Joule}}{m^3}$. 
The strain-energy density function relates  the amount of
energy that is stored within the material in response to a given
strain. Hence, the stresses in a hyperelastic material with a given
strain-energy density function, depends only on the strain, and not the
path for which the material deforms. On the contrary, if the model had
been visco-elastic we would expect to see hysteresis in the
stress/strain curve, but this is not possible for a hyperelastic
material.

\begin{remark}
  The second law of thermodynamics states that the total entropy
  production in a thermodynamic process can never be negative. Elastic
  materials defines a special class of materials in which the entropy
  production is zero. Within this thermodynamic framework the
  strain-energy density function coincides (up to a constant) with the
  Helmholtz free energy density.
\end{remark}

\subsubsection{General requirements for the strain-energy density
  function}
\label{sec:strain_energy_req}
Some general requirements must hold for the strain-energy function.
First of all, we require that the reference state is stress free and
that the stored energy increases monotonically with the deformation. 
Formally this can be stated simply as 
\begin{align*}
  \Psi(\I) = 0 \; && \text{and} &&\; \Psi(\F) \geq 0.
\end{align*}
Moreover, expanding or compressing a body to zero volume would
require an infinite amount of energy, i.e
\begin{align*}
  \Psi(\F) \rightarrow \infty \; && \text{as} &&\; \det \F \rightarrow& 0 \\
  \Psi(\F) \rightarrow \infty \; && \text{as} &&\; \det \F \rightarrow& \infty
\end{align*}
We say that the strain energy should be objective, meaning that the
stored energy in the material should be invariant with respect to
change of observer. Formally we must have: \emph{given any positive symmetric
rank-2 tensor $\C \in \mathrm{Sym}$:}
\begin{align}
  \Psi(\mathbf{C}) = \Psi(\mathbf{Q}\mathbf{C}\mathbf{Q}^T), \; \forall \mathbf{Q} \in \mathcal{G} \subseteq \mathrm{Orth}.
\end{align}
Here $\mathrm{Orth}$ is the group of all positive orthogonal matrices.
If $\mathcal{G} = \mathrm{Orth}$ we say that the material is
isotropic, and otherwise we say that the material is anisotropic.
This brings us to another important issue, which is related to the
choice of coordinate-system. Having to deal with different
coordinate-systems, and mapping quantities from one coordinate-system
to another can results in complicated computations. Therefore it would be beneficial if we
could work with quantities which do not depend on the choice of
coordinate-system. Such quantities are called invariants. 
If the material is isotropic, the representation theorem for
invariants states that $\Psi$ can be expressed in terms of the
principle invariants of $\mathbf{C}$, that is $\Psi = \Psi(I_1, I_2,
I_3)$. The principle invariants $I_i, i=1,2,3$ are the coefficients in
the characteristic polynomial of $\mathbf{C}$, and is given by 
\begin{align}
  I_1 = \tr \C,  && I_2 = \frac{1}{2}\left[ I_1^2 - \tr(\C^2)\right] && \text{and} && I_3 = \det \C.
\end{align}
In the case when the material constitutes a transversely isotropic
behavior, that is the material has a preferred direction $\mathbf{a}_0$,
which in the case of the myocardium could be the direction of fiber
muscle fibers, we have
\begin{align*}
  \mathcal{G} = \{ \mathbf{Q} \in \mathrm{Orth}: \mathbf{Q}(\mathbf{a}_0\otimes\mathbf{a}_0)\mathbf{Q}^T\},
\end{align*}
with $\otimes$ being the outer product. In this case the strain-energy
density function can still be expressed through invariants. However,
we need to include the so called quasi-invariants, which are defined
as stretches in the local microstructural coordinate-system. The
transversely isotropic invariants are given by
\begin{align*}
  I_{4\mathbf{a}_0 } = \mathbf{a}_0 \cdot (\C \mathbf{a}_0) && \text{and} && I_5 = \mathbf{a}_0 \cdot (\C^2 \mathbf{a}_0).
\end{align*}
The invariants do have a physical interpretation. For instance, $I_3$
is related to the volume ratio of material during deformation, while
$I_{4\mathbf{a}_0 } $ is related to the stretch along the direction
$\mathbf{a}_0 $. Indeed the \emph{stretch} ratio in the direction
$\mathbf{a}_0$ is given by $\lambda_{\mathbf{a}_0} = | \F \mathbf{a}_0
|$ and we see that $I_{4\mathbf{a}_0 }  =  \mathbf{a}_0 \cdot (\C
\mathbf{a}_0) = \F \mathbf{a}_0 \cdot (\F \mathbf{a}_0) =
\lambda_{\mathbf{a}_0}^2$. For more details about invariants see
\cite{holzapfel2009constitutive,liu1982representations}.

The theory of global existence of unique solutions for elastic problem
was originally based convexity of the free energy function.
A function $\phi: \Omega \rightarrow \mathbb{R}$ is \emph{convex} if for any
$\Xvec_1, \Xvec_2 \in \Omega$ we have
\begin{align}
  \phi(\lambda \Xvec_1+ (1-\lambda) \Xvec_2)
  \leq \lambda \phi(\Xvec_1)
  + (1-\lambda) \phi(\Xvec_2).
\end{align}
However, from a physical point of view this requirement is too strict
\cite{ball1976convexity}. A slightly weaker requirement is that the
strain-energy function $\Psi$ is \emph{polyconvex}, meaning that there exist
a convex function $\phi$ such that
\begin{align*}
  \Psi(\F) = \phi(\F, \cof \F, \det \F).
\end{align*}
Note however, that if $\Psi$ is convex then it is also polyconvex.

@@

The myocardium is composed by small blood vessels that supply the
myocardial cells with oxygen. When the myocardium contracts, this
perfused blood is squeezed out, resulting in an overall loss of
2-4\% volume\cite{yin1996compressibility}. Such materials are
referred to as compressible. When the 
volume are preserved we refer to the material as incompressible.
Since 2-4\% is very little, a common assumption in cardiac mechanical modeling,
which has also been made in the work conducted in this thesis, is to assume
the myocardium to be incompressible. The reason for this choice is
purely numerical.

For an incompressible material, only isochoric motions are
possible. This means that volume of the material do not change during
any deformation and hence we have the constraint 
\begin{align}
  J = \det(\F) = 1.
  \label{eq:incompressible_cons}
\end{align}
The constraint \eqref{eq:incompressible_cons} can be imposed by
considering the modified strain energy function
\begin{align}
  \Psi = \Psi(\F) + p(J-1),
  \label{eq:incomp_strain_energy}
\end{align}
where $p$ is a scalar which serves as a Lagrange multiplier, but which
can be identified as the hydrostatic pressure. If we differentiate
\eqref{eq:incomp_strain_energy} with respect to $\F$ we get the First
Piola-Kirchhoff stress tensor for an incompressible material
\begin{align}
  \FPK = \frac{\partial \Psi(\F)}{\partial \F} + J p \F^{-T}.
\end{align}
Likewise the Cauchy stress tensor is given by
\begin{align}
  \Cauchy = J^{-1} \frac{\partial \Psi(\F)}{\partial \F}\F^{T} + p \I.
  \label{eq:cauchy_incomp}
\end{align}

@@

\subsubsection{Uncoupling of volumetric and isochoric response}
The total strain energy function in \eqref{eq:incomp_strain_energy}
can be written as a sum of isochoric and volumetric components. Let
\begin{align}
  \F =  \F_{\mathrm{vol}} \F_{\mathrm{iso}},
\end{align}
then $ \F_{\mathrm{vol}} =
J^{1/3}\I$ and $\F_{\mathrm{iso}} = J^{-1/3}\F$. For
compressible material (i.e with $J \neq 1$) it is important to consider
only deviatoric strains in the strain-energy density function, so that
$\Psi = \Psi_{\mathrm{iso}}(\F_{\mathrm{iso}}) +
\Psi_{\mathrm{vol}}(J)$. For incompressible material ($J = 1$), we
have $\F_{\mathrm{vol}} = \I$ so that such a decomposition seems
unnecessary. However, a similar decomposition has shown to be
numerically beneficial \cite{weiss1996finite}. Note that, in this case, a similar
decoupling of the stress tensors has to be done.

@@

Choosing the correct boundary conditions for the model is essential,
and the choice should mimic what is observed in the reality. To
physiologically constrain the ventricle in a correct way is difficult,
and different approaches has been proposed.
The boundary condition at the endocardium is typically modeled as a
Neumann boundary condition, representing the endocardial blood
pressure. For the left ventricle we have
\begin{align}
  \sigma \mathbf{n} = -p_{\mathrm{lv}} \mathbf{n}, \;  \xvec \in  \lVendoCur, 
\end{align}
and for the right ventricle, ``lv'' is substituted with ``rv''.
This condition has a negative sign because the unit normal
$\mathbf{N}$ is pointing out of the domain, while the pressure is
acting into the domain. 
Note that this condition is imposed on the current configuration, and
to utilize the Lagrangian formulation we can pull back this condition
to the reference configuration to obtain
\begin{align}
  \FPK\mathbf{N} &= -p_{\mathrm{lv}} J\mathbf{F}^{-T} \cdot \mathbf{N}, \;  \Xvec \in \lVendo
\end{align}
Likewise, it is common to enforce a
Neumann boundary condition on the epicardium,
\begin{align}
\FPK\mathbf{N}  &= -p_{\mathrm{epi}}  J\mathbf{F}^{-T} \cdot \mathbf{N}, \;  \Xvec \in \epi.
\end{align}
However, the pressure $p_{\mathrm{epi}}$ is usually set
to zero.

@@

\subsection{Force-balance equation}
We will now collect all the terms that makes up to force balance for
the cardiac mechanics problem. Considering the myocardium as a incompressible,
hyperelastic material with a Robin type boundary condition at the base
we obtain the following strong form in the Lagrangian formulation
\begin{align}
  \begin{split}
  \nabla \cdot \FPK &= 0 \\
  J - 1 &= 0,
  \end{split}
 \label{eq:force_balance_strong}
\end{align}
completed with appropriate boundary conditions. To solve this
numerically using the finite element method, we need to derrive weak
variational form of this equaiton.
% \begin{align}
%   \FPK \Nvec + k \uvec &= \mathbf{0},  \;  \Xvec \in \lvbase\\
%   u_1 &= 0,  \;  \Xvec \in \lvbase \\
%   \FPK\mathbf{N}  &= \mathbf{0}, \;  \Xvec \in \epi \\
%   \FPK\mathbf{N} &= -p_{\mathrm{lv}} J\mathbf{F}^{-T} \cdot \mathbf{N}, \;  \Xvec \in \lvendo
%   \label{eq:bndry_conditions_strong}
% \end{align}

@@

  \Pi(\uvec, p) &= \Pi_{\mathrm{int}}(\mathbf{u},p) + \Pi_{\mathrm{ext}}(\mathbf{u}). \\
  \Pi_{\mathrm{int}}(\uvec,p) &= \int_{\Omega} \left[ p(J - 1) +  \Psi(\mathbf{F}) \right] \mathrm{d}V\\

@@

 \label{eq:intro_variational_form}
